---
title: 'MIE237'
author: "Neil Montgomery"
date: "2016-03-01"
output: 
  ioslides_presentation: 
    css: 'styles.css' 
    widescreen: true 
    transition: 0.001
---
\newcommand{\Var}[1]{\text{Var}\left( #1 \right)}
\newcommand{\E}[1]{E\left( #1 \right)}
\newcommand{\Sample}[1]{#1_1,\ldots,#1_n}
\newcommand{\od}[2]{\overline #1_{#2\cdot}}
\newcommand{\flist}[2]{\{#1_1, #1_2, \ldots, #1_#2\}}
\newcommand{\samp}[2]{#1_1, #1_2, \ldots, #1_#2}
\renewcommand{\bar}[1]{\overline{#1}}
\newcommand{\ve}{\varepsilon}
\newcommand{\bs}[1]{\boldsymbol{#1}}



```{r, echo=FALSE}
knitr::opts_chunk$set(echo=FALSE)
```



```{r, message=FALSE}
library(dplyr)
library(ggplot2)
set.seed(2)
n <- 40
x <- seq(1, 10, length.out = n)
y1 <-  1 + x + rnorm(n, 0, 0.1)

y2 <- 1 - x + rnorm(n, 0, 1)

y3 <- 1 + x + rnorm(n, 0, 5)

set.seed(5)
y4 <- 1 + x + rnorm(n, 0, 2)

y5 <- 20 - (x - 8)^2 + rnorm(n, 0, 8)


set.seed(42)
y6 <- 1 + x + rnorm(n, 0, 2)

x2 <- seq(1, 10, length.out = 2*n)
y7 <- 1 + x2 + rnorm(2*n, 0, 2)
```

# New Topic: Correlation

## Preview of "correlation" (from 2016-01-29)

Recall $\text{Cov}(X,Y) = E(X-E(X))(Y-E(Y)) = E(XY)-E(X)E(Y)$.

Measures linear relationship between $X$ and $Y$.

A related measure that divides by the two standard deviations:
$$\rho = \frac{\text{Cov}(X,Y)}{\sqrt{\Var X \Var Y}}$$

"Correlation coefficient" $\rho$ is a \textit{theoretical} measure of the linear relationship between the random variables $X$ and $Y$. We'd like an analogous *empirical* measure for *sample data*.

## Correlation

```{r, echo=FALSE, fig.height=3, fig.width=3}
data.frame(x, y1) %>% 
  ggplot(aes(x=x, y=y1)) + geom_point() +
  geom_hline(yintercept = mean(y1)) + 
  geom_vline(xintercept = mean(x))

data.frame(x, y2) %>% 
  ggplot(aes(x=x, y=y2)) + geom_point() +
  geom_hline(yintercept = mean(y2)) + 
  geom_vline(xintercept = mean(x))
```

'Twill be based on:
$$S_{xy}=\sum_{i=1}^n \left(x_i-\overline x\right)
\left(y_i - \overline y\right)$$

## Sample correlation coefficient { .build }

Define:
$$r = \frac{S_{xy}}{\sqrt{S_{xx}{S_{yy}}}}$$

$r$ is an estimator of $\rho$, called the "sample correlation coefficient". 

In regression analysis the $x$ variable is considered fixed, and $y$ is a function of $x$.

Correlation is a property of $Y$ and $X$ both random, and correlation is symmetric in $X$ and $Y$.

Nevertheless we use (misuse?) sample correlation coefficient also in a regression context as a useful numerical summary of a bivariate relationship.

## Connection with $R^2$

Recall that $S_{yy} = \sum (y_i - \bar y)^2 = SST$ and $R^2 = SSR\big/SST = 1-SSE\big/SST$.

It turns out also that $SSE = S_{yy} - \hat\beta_1S_{xy}$. So:
$$\begin{align*}
R^2 &= 1 - \frac{SSE}{SST}\\
&= 1 - \frac{S_{yy} - \hat\beta_1S_{xy}}{S_{yy}}\\
&= 1 - \left(1 - \frac{\hat\beta_1S_{xy}}{S_{yy}}\right)\\
&= \frac{S_{xy}^2}{S_{xx}S_{yy}} = (r)^2\end{align*}$$

## Other properties of $r$ { .build }

Since $0 \le R^2 \le 1$ it follows $-1 \le r \le 1$.

Also:
$$r = \hat\beta_1\sqrt{\frac{S_{xy}}{S_{xx}}}$$

It is possible to do things like test $H_0: \rho=0$ versus $H_1: \rho\ne 0$ using $r$, but we will not cover this.

## Examples

```{r, fig.width=3, fig.height=2.5}
data.frame(x, y1) %>% 
  ggplot(aes(x=x, y=y1)) + geom_point() + 
  ggtitle(substitute(paste(r, " = ", r1), list(r1 = round(cor(x, y1), 3))))
data.frame(x, y2) %>% 
  ggplot(aes(x=x, y=y2)) + geom_point() + 
  ggtitle(substitute(paste(r, " = ", r2), list(r2 = round(cor(x, y2), 3))))
data.frame(x, y3) %>% 
  ggplot(aes(x=x, y=y3)) + geom_point() + 
  ggtitle(substitute(paste(r, " = ", r3), list(r3 = round(cor(x, y3), 3))))

y9 <- 10 + rnorm(n, 0, 3)
data.frame(x, y9) %>% 
  ggplot(aes(x=x, y=y9)) + geom_point() + 
  ggtitle(substitute(paste(r, " = ", r9), list(r9 = round(cor(x, y9), 3))))


y10 <- 10 + (x - mean(x))^2 + rnorm(n, 0, 0.01)
data.frame(x, y10) %>% 
  ggplot(aes(x=x, y=y10)) + geom_point() + 
  ggtitle(substitute(paste(r, " = ", r10), list(r10 = round(cor(x, y10), 3))))



x3 <- c(x, max(x) + 12.5)
y11 <- c(rnorm(n, 0, 0.25), 3)
data.frame(x3, y11) %>% 
  ggplot(aes(x=x3, y=y11)) + geom_point() + 
  ggtitle(substitute(paste(r, " = ", r11), list(r11 = round(cor(x3, y11), 3))))
```


